Reproducible methods for modeling combined public transport and cycling trips and associated benefits: evidence from the biclaR tool

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.Rmd
maps pop and trips

library(dplyr)
library(ggplot2)
```

@@

According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.
Car modal share was 58.4%, while PT accounted for 15.5% (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
This exceeds the number of inter-municipal trips (1.8 million trips), involving travel between different municipalities.
Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
`r round(100*2948314/5569731)`% trips are up to 5 km distance, and `r round(100*3958667/5569731)`% up to 10 km.
Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.

@@

**TO-DO**

-   mais mapas:
-   Densidade populational - nº de viagens
-   Pares OD - desire lines width

@@

Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we disaggregated the OD data into desire lines reprenting a maximum of 100 trips each.
Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.

@@

MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
stopsAML = sf::st_read("load/Stops_AML.geojson", quiet = TRUE)
stopsAML = stopsAML |> filter(mode != "Metropolitano") |>
  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
                       labels = c("Train", "Ferry", "Tram", "Bus")))
routesAML = sf::st_read("load/Routes_AML.geojson", quiet = TRUE)
routesAML = routesAML |> 
  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
                       labels = c("Train", "Ferry", "Tram", "Bus")))

>>>>

library(dplyr)
library(ggplot2)
library(sf)
```

@@

The Lisbon metropolitan area (LMA), in Portugal, is the case study, comprising 18 municipalities, with a total area of 3,015 km² and a population of 2.8 million residents
The LMA is the most populous and economically active region in Portugal, with the city of Lisbon as the central node of the LMA, with a high concentration of jobs and services, and a high demand for mobility (see Figure \ref{fig:lma}).
As shown, the number of trips is proportional to the population of each municipality, with the highest number of trips originating in Lisbon, followed by the municipalities of Sintra, Cascais, Loures, and Almada.

The Public Transport network is well developed, although very much Lisbon-centric, with a high level of service, including inter-municipal trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a metro system that serves Lisbon and surrounding municipalities.

```{r lma, fig.cap="\\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.", fig.subcap=c("Population in the LMA","Trips/day in the LMA"), fig.ncol=2, message=FALSE, warning=FALSE, out.width="50%"}
MUNICIPIOSgeo = readRDS("load/MUNICIPIOSgeo.Rds")
MUNICIPIOSgeo_mapa = readRDS("load/MUNICIPIOSgeo_mapa.Rds")
MUNICIPIOS_dtcc = readRDS("load/MUNICIPIOS_dtcc.Rds")
POPMunicipios = readRDS("load/POPMunicipios.Rds")

# labels
cities = MUNICIPIOSgeo |> 
  filter(Concelho %in% c("Lisboa", "Sintra", "Cascais", "Loures", "Almada"))
cities$point = sf::st_centroid(cities$geometry)
cities = sf::st_set_geometry(cities, cities$point)

# population
pop_mun = POPMunicipios |> 
  left_join(MUNICIPIOS_dtcc, by = c("DTCC" = "DTMN"))
pop_mun$Concelho[pop_mun$Concelho == "Setúbal"] = "Setubal"
pop_mun = MUNICIPIOSgeo |>
  left_join(pop_mun, by = "Concelho")
pop_mun$area = sf::st_area(pop_mun) |> units::drop_units()
pop_mun$popdens = pop_mun$Populacao / pop_mun$area * 1000000

# residents
  ggplot() +
  geom_sf(data = pop_mun,
          aes(fill = Populacao),
          color = "gray60") +
  scale_fill_gradient(low = "white", high = "darkred",
                      labels = scales::unit_format(unit = "k", scale = 1e-3)
                      ) +
  theme_void() +
  # theme(legend.position = "bottom")+
  labs(fill = "Residents") +
  geom_sf_text(data = cities,
               aes(label = Concelho),
               size = 2.3)

# #denstity
# ggplot() +
#   geom_sf(data = pop_mun,
#           aes(fill = popdens),
#           color = "gray60") +
#   scale_fill_gradient(low = "white", high = "darkred") +
#   theme_void() +
#   labs(fill = "inhabitants/km²")

# trips
ggplot() +
  geom_sf(data = MUNICIPIOSgeo_mapa,
          aes(fill = viagens),
          color = "gray60") +
  viridis::scale_fill_viridis(option = "F", direction = -1,
                              labels = scales::unit_format(unit = "k", scale = 1e-3)) +
  theme_void() +
  labs(fill = "Trips / day")

The LMA is a low cycling maturity area, characterised by a sparse and fragmented cycling network totalling 378 km, primarily concentrated within Lisbon, and a low cycling modal share.
According to the latest mobility survey conducted in 2018 [@IMOB], the LMA registered a total of 5.3 million daily trips, with only 0.5% by bicycle.

The survey also showed that 58.4% of trips were made by car, 15.5% by public transport, and 23.9% by walking (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in the same municipality --- amounts to 3.5 million trips.
This exceeds the number of inter-municipal trips (1.8 million trips), involving travel between different municipalities.
Cars and public transport are the most used modes for intercity trips, with cars being the predominant choice for all journeys.
`r round(100*2948314/5569731)`% trips are up to 5 km distance, and `r round(100*3958667/5569731)`% up to 10 km.
Nevertheless, 29% of trips are longer than 10 km, which requires the use of motorized modes, or active modes in combination with public transport.

@@

@@

Using the [`odjitter` R package](https://github.com/dabreegster/odjitter), we disaggregated the OD data into desire lines reprenting a maximum of 100 trips each.
Figure \ref{fig:jitter} illustrates the contrast between trip representation through the traditional method, which connects a single desire line between each district, and the presentation achieved through the randomization and disaggregation of trips between districts, specifically for the Lisbon metropolitan area.
As shown, the city of Lisbon (in the centre) is the main attractor of trips, with a high number of trips to and from the other municipalities.

@@

stopsAML = sf::st_read("load/Stops_AML.geojson", quiet = TRUE)
stopsAML = stopsAML |> filter(mode != "Metropolitano") |>
  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
                       labels = c("Train", "Ferry", "Tram", "Bus")))
routesAML = sf::st_read("load/Routes_AML.geojson", quiet = TRUE)
routesAML = routesAML |> 
  mutate(mode = factor(mode, levels = c("Comboio", "Ferry", "Metro Superficie", "Autocarro"),
                       labels = c("Train", "Ferry", "Tram", "Bus")))

### 2024-05-08 paper/PaperCEUS/PaperCEUS_rev1.tex
maps pop and trips

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326, van2021insights}. In The Netherlands, which
has the highest mode share of cycling in the world, cycling accounts for
more than a third of all trips to and from rail stations at the `home'
end of the journey, greatly increasing the ability of the transport
system \citep{RIETVELD200071}. This approach to reducing car dependency
and associated externalities requires interventions and programs to make
bicycling more appealing \citep{lapaix_role_2021}. The resulting public
investments can have significant social and environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

@@

According to the latest mobility survey conducted in 2018 \citep{IMOB},
the LMA registered a total of 5.3 million daily trips, with only 0.5\%
by bicycle. Car modal share was 58.4\%, while PT accounted for 15.5\%
(see Figure \ref{fig:mododist}). The number of intra-municipal trips ---
with origin and destination in the same municipality --- amounts to 3.5
million trips. This exceeds the number of inter-municipal trips (1.8
million trips), involving travel between different municipalities. Cars
and public transport are the most used modes for intercity trips, with
cars being the predominant choice for all journeys. 53\% trips are up to
5 km distance, and 71\% up to 10 km. Nevertheless, 29\% of trips are
longer than 10 km, which requires the use of motorized modes, or active
modes in combination with public transport.

@@

\textbf{TO-DO}

\begin{itemize}
\tightlist
\item
  mais mapas:
\item
  Densidade populational - nº de viagens
\item
  Pares OD - desire lines width
\end{itemize}

@@

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we disaggregated the OD data into desire lines reprenting a
maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.

>>>>

Combining public transportation (PT) and cycling for the first and last
mile in metropolitan areas can replace a high proportion of private car
trips \citep{MARTENS2007326, vanmil_insights_2021}. In The Netherlands,
which has the highest mode share of cycling in the world, cycling
accounts for more than a third of all trips to and from rail stations at
the `home' end of the journey, greatly increasing the ability of the
transport system \citep{RIETVELD200071}. This approach to reducing car
dependency and associated externalities requires interventions and
programs to make bicycling more appealing \citep{lapaix_role_2021}. The
resulting public investments can have significant social and
environmental benefits
\citep{internationaltransportforum_integrating_2017}. Despite the
benefits of cycling-PT intermodality, the potential of this combination
is often overlooked in transport planning \citep{lapaix_role_2021}.

@@

The Lisbon metropolitan area (LMA), in Portugal, is the case study,
comprising 18 municipalities, with a total area of 3,015 km² and a
population of 2.8 million residents The LMA is the most populous and
economically active region in Portugal, with the city of Lisbon as the
central node of the LMA, with a high concentration of jobs and services,
and a high demand for mobility (see Figure \ref{fig:lma}). As shown, the
number of trips is proportional to the population of each municipality,
with the highest number of trips originating in Lisbon, followed by the
municipalities of Sintra, Cascais, Loures, and Almada.

The Public Transport network is well developed, although very much
Lisbon-centric, with a high level of service, including inter-municipal
trains, buses, trams, and ferries (see Figure \ref{fig:map1}), and a
metro system that serves Lisbon and surrounding municipalities.

\begin{figure}
\subfloat[Population in the LMA\label{fig:lma-1}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-1} }\subfloat[Trips/day in the LMA\label{fig:lma-2}]{\includegraphics[width=0.5\linewidth,]{PaperCEUS_rev1_files/figure-latex/lma-2} }\caption{\label{lma}Population and daily trips (origins) at each Municipality of the LMA according to census and the travel survey.}\label{fig:lma}
\end{figure}

The LMA is a low cycling maturity area, characterised by a sparse and
fragmented cycling network totalling 378 km, primarily concentrated
within Lisbon, and a low cycling modal share. According to the latest
mobility survey conducted in 2018 \citep{IMOB}, the LMA registered a
total of 5.3 million daily trips, with only 0.5\% by bicycle.

The survey also showed that 58.4\% of trips were made by car, 15.5\% by
public transport, and 23.9\% by walking (see Figure \ref{fig:mododist}).
The number of intra-municipal trips --- with origin and destination in
the same municipality --- amounts to 3.5 million trips. This exceeds the
number of inter-municipal trips (1.8 million trips), involving travel
between different municipalities. Cars and public transport are the most
used modes for intercity trips, with cars being the predominant choice
for all journeys. 53\% trips are up to 5 km distance, and 71\% up to 10
km. Nevertheless, 29\% of trips are longer than 10 km, which requires
the use of motorized modes, or active modes in combination with public
transport.

@@

@@

Using the
\href{https://github.com/dabreegster/odjitter}{\texttt{odjitter} R
package}, we disaggregated the OD data into desire lines reprenting a
maximum of 100 trips each. Figure \ref{fig:jitter} illustrates the
contrast between trip representation through the traditional method,
which connects a single desire line between each district, and the
presentation achieved through the randomization and disaggregation of
trips between districts, specifically for the Lisbon metropolitan area.
As shown, the city of Lisbon (in the centre) is the main attractor of
trips, with a high number of trips to and from the other municipalities.
